```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
```
## Day 1: Tidying data
### In-class worksheet

**June 21st, 2021**

In this worksheet, we will use the library tidyverse:
```{r message=FALSE}
library(tidyverse)
```

## Tidying data with `pivot_longer()`and `pivot_wider()`

### 1.1 Making tables longer

Consider the following two data sets, `male_haireyecolor` and `female_haireyecolor`. The data sets record the occurrence of hair and eye color phenotype combinations in a class of statistics students. Use `head()` to preview these data sets; are they tidy?

```{r message = FALSE}
# download male data set
male_haireyecolor <- read_csv("https://rachaelcox.github.io/classes/datasets/male_haireyecolor.csv")

# download female data set
female_haireyecolor <- read_csv("https://rachaelcox.github.io/classes/datasets/female_haireyecolor.csv")
```

Use the function `pivot_longer()` to rearrange both data sets such that there is one observation per row for each combination of hair and eye color. **Remember:** You can run `?pivot_longer` to pull up argument details and example usage.

```{r}
# your R code here
```


### 1.2 Making tables wider

Consider the following data set `persons`, which contains information about the sex, weight, and height of 200 individuals. Use `head()` to preview the data set; is it tidy?

```{r message = FALSE}
# download persons data set
persons <- read_csv("https://rachaelcox.github.io/classes/datasets/persons.csv")
```

Rearrange the `persons` data frame so that you have one column for subject, one for sex, one for weight, and one for height using the function `pivot_wider`. **Remember:** You can run `?pivot_wider` to pull up argument details and example usage.

```{r}
# your R code here
```
