```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
```
## Day 1: Tidying data
### In-class worksheet, solutions
**June 29, 2020**

In this worksheet, we will use the library tidyverse:
```{r message=FALSE}
library(tidyverse)
```

## Tidying data (`pivot_longer()`, `pivot_wider()`)

### 1.1 Making tables longer

Consider the following two data sets, `male_haireyecolor` and `female_haireyecolor`. The data sets record the occurrence of hair and eye color phenotype combinations in a class of statistics students. Use `head()` to preview these data sets; are they tidy?

```{r message = FALSE}
# download male data set
male_haireyecolor <- read_csv("https://rachaelcox.github.io/classes/datasets/male_haireyecolor.csv")
head(male_haireyecolor)

# download female data set
female_haireyecolor <- read_csv("https://rachaelcox.github.io/classes/datasets/female_haireyecolor.csv")
head(female_haireyecolor)
```

The data set is not tidy, because the columns `Brown`, `Blue`, `Hazel`, and `Green` are observations for a single variable, eye color. The following versions of the tables are tidy:

```{r}
# pivot all columns except the column `Hair`
male_colors_tidy <- male_haireyecolor %>%
  pivot_longer(-Hair, names_to = "EyeColor", values_to = "Count")
head(male_colors_tidy)

# pivot all columns except the column `Hair`
female_colors_tidy <- female_haireyecolor %>%
  pivot_longer(-Hair, names_to = "EyeColor", values_to = "Count")
head(female_colors_tidy)
```

There is another syntax that specifies a range of columns that also works:

```{r}
# pivot all columns in the range of `Brown` to `Green`
male_colors_tidy <- male_haireyecolor %>%
  pivot_longer(Brown:Green, names_to = "EyeColor", values_to = "Count")
head(male_colors_tidy)

# pivot all columns except the column `Hair`
female_colors_tidy <- female_haireyecolor %>%
  pivot_longer(Brown:Green, names_to = "EyeColor", values_to = "Count")
head(female_colors_tidy)
```

### 1.2 Making tables wider

Consider the following data set `persons`, which contains information about the sex, weight, and height of 200 individuals:

```{r message = FALSE}
persons <- read_csv("https://rachaelcox.github.io/classes/datasets/persons.csv")
head(persons)
```

Is this data set tidy? And can you rearrange it so that you have one column for subject, one for sex, one for weight, and one for height?

The data set is not tidy, because neither `indicator` nor `value` are variables. The variables are `subject`, `sex`, `weight`, `height`. The following version of the table is tidy:

```{r}
persons %>%
  pivot_wider(names_from = "indicator", values_from = "value") %>%
  head()
```
